Hazardous asteroids forecast via Markov random fields

### 2021-12-16 OLD/Manuscript/manuscript.tex
PUB

\usepackage{braket}
\usepackage{changepage}
\usepackage[english]{babel}	% latest language is predefined
\usepackage[T1]{fontenc}		% font coding

@@

In this section, I am going to review the theoretical concepts
that underlie the probabilistic methods here used: these will be
exposed following the approaches of Murphy \cite{murphy2012machine} , Koller
et al. \citep{koller2009probabilistic}, Højsgaardand et al. \citep{hojsgaard2012graphical} and Russel et al. \citep{russell2010artificial}. Furthermore, a rapid overview of the main concepts of information theory will also be provided following the Cover \citep{cover2006elements}, and MacKay \citep{mackay2003information} approaches. This is because some of its quantities in the preliminary analysis of the dataset were used. Finally, I concluded this chapter with a rapid overview of algorithm interpretability, which, together with the accuracy and the
confusion matrix, is necessary for the assessment of a ML performances. On the other side, the concepts related to the celestial
mechanics here used will be described, following the Murray
approach \cite{murray1999solar} into the Appendix A.

@@

Where $w_{st}=J$ represent the coupling strength between two neighbour site.  The collective state is described by

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Perihelion_Distance.pdf}
    \subcaption{ \begin{center}
    a) Perihelion Distance
    \end{center}}
    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Eccentricity.pdf}
    \subcaption{ \begin{center}
    b) Eccentricity
    \end{center}}
    \vspace{4ex}
  \end{minipage} \\
    \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Min_orbit_intersection.pdf}
    \subcaption{ \begin{center}
    c) Min orbit intersection
    \end{center}}
    \vspace{4ex}
  \end{minipage}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Absolute_magnitude.pdf}
    \subcaption{ \begin{center}
    d) Absolute magnitude
    \end{center}}
    \vspace{4ex}
  \end{minipage}
\caption{Comparison between the density distributions of hazardous (red) and non-hazardous (light blue) asteroids for a selected set of features that, according to the theory, are interesting.}
\label{Density_relevant}
\end{figure}

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are directly connected to the Hazardous feature are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the relative velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Moving to the relation of orbital parameters, it can be seen that there is a negative relation between the eccentricity and perihelion distance: this is meaningful since it comes from the definition of eccentricity. The other side of the coin of this connection is the positive relationships within the semimajor axis and the eccentricity. Furthermore, the positive relationship between the mean motion and the semimajor axis is correct: this is the result of the 3$^{th}$ Kepler law. Finally the connections with the Jupiter Tisserand Invariant comes directly from the definition of this parameter. A full explanation of these connections is provided in the Appendix A. Thus it can be said that the model produced is consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC (Receiver operating characteristic) curve and $\phi$ coefficient (also known as Matthews correlation coefficient ) were evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where the result is obtained without boundaries should be preferred to those obtained with a large number of boundaries. Therefore we reject the mmod and minforest ones favouring the mgm, which was obtained without boundaries.

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[width=1.5\textheight]{Figures/mmod.pdf}
\caption{The graphical mixed interaction model obtained with the grim package \cite{hojsgaard2012graphical}. The plot was obtained with the igraph package for R \cite{igraph} }
\label{mmod}
\end{center}
\end{figure}
\end{landscape}
\\
\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[width=1.5\textheight]{Figures/minforest.pdf}
\caption{The graphical mixed interaction model obtained with the gRapHD package \cite{de2009high} with a stepwise algorithm. The plot was obtained with the \cite{qgraph} package }
\label{minforest}
\end{center}
\end{figure}
\end{landscape}

@@

\section{Machine learning algorithms} The performances of mgm model will be now compared with the one obtained from other ML algorithms. These include the random-forest (RF, as implemented in the RandomForest package \cite{rfor}), the support vector machines (SVM, as implemented in the e1071 package \cite{dimitriadou2008misc}), the quadratic discriminant anchiedersialysis (QDA, as implemented in the MASS package \cite{MASin reverseS} and the logistic regression (as implemented in stats package \cite{stats}). Their performances are reported in Fig. \ref{CF_ML} and \ref{ROC_mgm} as well in the Tab. \ref{phi_values}. This comparison shows that the RF and the SVM outperform the mgm graphical method while the logistic one has similar performances. Thus, one can wonder about the advantage of using a graphical method instead of a random forest or an SVM since its performances seem lower. The answer is the interpretability of the model provided: the RF, at least, can provide variable rank importance as reported in Fig. \ref{RF_Importance} (note that apart from the Min Orbit intersection, there is a slight reshuffling in the feature importance to the one estain reverseblished by the MI in fig \ref{Mutual_information}. However, apart from this intepretation, the RF is a black-box as the SVM, the logistic regression and the QDA. Conversely, the probabilistic graphs, providing the list of connections among the random variables, give the user an interpretable model whose properties can also be compared, discussed, and validated with the theory. Thus the model developed in this way allows a more scientific evaluation with respect to the black-box ones. In the author's view, this characteristic compensates for the lack of predictive power regarding the RF or the SVM methods.

@@

The GLASSO (with a penalizing parameter $\rho=0.3$) and the mgm algorithms provided graphical models that correctly reproduce the physical connections of the dataset feature. In particular, for both models, it can be seen that no one of the orbital parameters is conneted to the diameter of the asteroid, as stated by theory. Furthermore, the mgm model can provide forecasts about the hazardousness of the asteroids with almost equal or better performances than the logistic and QDA algorithms.  On the other hand, despite the mgm performances being lower than the RF and SVM, its interpretability is higher.  Such a feature compensates the mgm forecast performances gap. As a future outlook I consider to extend this analysis for physical phenomena for which there is no deterministic theory but only probabilistic one\footnote{Following the I.Prigogine \cite{prigogine2017non,prigogine1997end,nicolis1989exploring,prigogine1978time} argument it must be said that a pure deterministic theory, if exist, cannot be fully applied. Therefore, in principle, all possible theories should be considered probabilistic. This is because the initial conditions can not be known with a infinite precision. Therefore, if one also perfectly knows the equation of motion and can also solve them exactly (and in most of cases it is not possible) a source of error still persist and it propagates with the Lyapunov exponent. However as it was shown by Prigogine it a system is at equilibrium this instability can neglected. In this case a deterministic approach is meaningful }. For instance one can consider Earthquakes and Vulcanos Eruptions \cite{sornette1989self,sparks2003forecasting,sornette2006critical}

@@

\chapter{Appendix B: R code} Here I provide the code which I run to obtain the results previously showed.  For its writing I took in consideration the examples provided in the  packages documentation, the Prof.Nicolussi lectures and he examples provided in the Hojsgaard textbook  \cite{hojsgaard2012graphical}.

>>>>

\usepackage{braket}
\usepackage{changepage}
\usepackage[english]{babel}	% latest language is predefined
\usepackage[T1]{fontenc}		% font coding
\usepackage{pifont}

@@

In this section, I am going to review the theoretical concepts
that underlie the probabilistic methods here used: these will be
exposed following the approaches of Murphy \cite{murphy2012machine} , Kolleret al. \cite{koller2009probabilistic}, Højsgaardand et al. \cite{hojsgaard2012graphical} and Russel et al. \cite{russell2010artificial}. Furthermore, a rapid overview of the main concepts of information theory will also be provided following the Cover \cite{cover2006elements}, and MacKay \cite{mackay2003information} approaches. This is because some of its quantities in the preliminary analysis of the dataset were used. Finally, I concluded this chapter with a rapid overview of algorithm interpretability, which, together with the accuracy and the
confusion matrix, is necessary for the assessment of a ML performances. On the other side, the concepts related to the celestial
mechanics here used will be described, following the Murray
approach \cite{murray1999solar} into the Appendix A.

@@

Where $w_{st}=J$ represent the coupling strength between two neighbour site.  The collective state is described by \cite{murphy2012machine}:

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Perihelion_Distance.pdf}
    \subcaption{ \begin{center}
    a) Perihelion Distance
    \end{center}}
    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Eccentricity.pdf}
    \subcaption{ \begin{center}
    b) Eccentricity
    \end{center}}
    \vspace{4ex}
  \end{minipage} \\
    \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Min_orbit_intersection.pdf}
    \subcaption{ \begin{center}
    c) Min orbit intersection
    \end{center}}
    \vspace{4ex}
  \end{minipage}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{Figures/DENSITY_Absolute_magnitude.pdf}
    \subcaption{ \begin{center}
    d) Absolute magnitude
    \end{center}}
    \vspace{4ex}
  \end{minipage}
\caption{Comparison between the density distributions of hazardous (red) and non-hazardous (light blue) asteroids for a selected set of features that, according to the theory, are interesting. Plot obtained with the ggplot2 package \cite{ggplot2}}
\label{Density_relevant}
\end{figure}

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are directly connected to the Hazardous feature are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the relative velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Moving to the relation of orbital parameters, it can be seen that there is a negative relation between the eccentricity and perihelion distance: this is meaningful since it comes from the definition of eccentricity. The other side of the coin of this connection is the positive relationships within the semimajor axis and the eccentricity. Finally the connections with the Jupiter Tisserand Invariant comes directly from the definition of this parameter. Such arguments can be much more clear with the equations and plots provided in Appendix A. Thus it can be said that the model produced is almost consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC (Receiver operating characteristic) curve and $\phi$ coefficient (also known as Matthews correlation coefficient ) were evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where a better\footnote{In terms of consistency with astronomic laws} result is obtained without boundaries should be preferred to those obtained with a large number of constraints. Therefore we reject the mmod and minforest ones favouring the mgm, obtained without constraints (blacklist and/or whitelist) .

@@

\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[width=1.5\textheight]{Figures/mmod.pdf}
\caption{The graphical mixed interaction model obtained with the grim package \cite{hojsgaard2012graphical}. The plot was obtained with the igraph package for R \cite{igraph} }
\label{mmod}
\end{center}
\end{figure}
\end{landscape}
\\
\begin{landscape}
\begin{figure}
\begin{center}
\includegraphics[width=1.5\textheight]{Figures/minforest.pdf}
\caption{The graphical mixed interaction model obtained with the gRapHD package \cite{de2009high} with a stepwise algorithm. The plot was obtained with the qgraph \cite{qgraph} package }
\label{minforest}
\end{center}
\end{figure}
\end{landscape}

@@

\section{Machine learning algorithms} The performances of mgm model will be now compared with the one obtained from other ML algorithms. These include the random-forest (RF, as implemented in the RandomForest package \cite{rfor}), the support vector machines (SVM, as implemented in the e1071 package \cite{dimitriadou2008misc}), the quadratic discriminant analysis (QDA, as implemented in the MASS package \cite{MASS} and the logistic regression (as implemented in stats package \cite{stats}). Their performances are reported in Fig. \ref{CF_ML} and \ref{ROC_mgm} as well in the Tab. \ref{phi_values}. This comparison shows that the RF and the SVM outperform the mgm graphical method while the logistic one has similar performances. Thus, one can wonder about the advantage of using a graphical method instead of a random forest or an SVM since its performances seem lower. The answer is the interpretability of the model provided: the RF, at least, can provide variable rank importance as reported in Fig. \ref{RF_Importance} (note that apart from the Min Orbit intersection, there is a slight reshuffling in the feature importance to the one estain reverseblished by the MI in fig \ref{Mutual_information}. However, apart from this intepretation, the RF is a black-box as the SVM, the logistic regression and the QDA. Conversely, the probabilistic graphs, providing the list of connections among the random variables, give the user an interpretable model whose properties can also be compared, discussed, and validated with the theory. Thus the model developed in this way allows a more scientific evaluation with respect to the black-box ones. In the author's view, this characteristic compensates for the lack of predictive power regarding the RF or the SVM methods.

@@

The GLASSO (with a penalizing parameter $\rho=0.3$) and the mgm algorithms provided graphical models that almost correctly reproduce the physical connections of the dataset feature. In particular, for both models, it can be seen that no one of the orbital parameters is conneted to the diameter of the asteroid, as stated by theory. Furthermore, the mgm model can provide forecasts about the hazardousness of the asteroids with almost equal or better performances than the logistic and QDA algorithms.  On the other hand, despite the mgm performances being lower than the RF and SVM, its interpretability is higher.  Such a feature compensates the mgm forecast performances gap. As a future outlook I consider to extend this analysis for physical phenomena for which there is no deterministic theory but only probabilistic one\footnote{Following the I.Prigogine \cite{prigogine2017non,prigogine1997end,nicolis1989exploring,prigogine1978time} argument it must be said that a pure deterministic theory, if exist, cannot be fully applied. Therefore, in principle, all possible theories should be considered probabilistic. This is because the initial conditions can not be known with a infinite precision. Therefore, if one also perfectly knows the equation of motion and can also solve them exactly (and in most of cases it is not possible) a source of error still persist and it propagates with the Lyapunov exponent. However as it was shown by Prigogine it a system is at equilibrium this instability can neglected. In this case a deterministic approach is meaningful }. For instance one can consider Earthquakes and Vulcanos Eruptions \cite{sornette1989self,sparks2003forecasting,sornette2006critical}

@@

\chapter{Appendix B: R code} Here I provide the code which I run to obtain the results previously showed.  For its writing I took in consideration the examples provided in the  packages documentation, the Prof.Nicolussi lectures and the examples provided in the Hojsgaard textbook  \cite{hojsgaard2012graphical}.

### 2021-12-16 Presentation/Presentation.tex
PUB

% Uncomment these lines for an automatically generated outline.
\begin{frame}{Outline}
\setcounter{tocdepth}{1}
\begin{center}
  \tableofcontents
\end{center}
\end{frame}

@@

\begin{frame}{Celestial mechanics \cite{murray1999solar}: Magnitude}
\begin{equation}
\Phi=\frac{L}{4\pi r^{2}}
\end{equation}
\begin{equation}
m=-2.5\log_{10}\Phi+C
\end{equation}
\begin{equation}
m_{1}-m_{2}=-2.5\log_{10}\frac{\Phi_{1}}{\Phi_{2}}
\end{equation}
\begin{equation}
M-m=-2.5\log_{10}\frac{\Phi\cdot d^{2} }{\Phi\cdot 10^{2}}
\end{equation}
\begin{equation}
M=m+5-5log_{10}d
\end{equation}
Where $\Phi$ is the flux for a sphere of radius r, m the relative magnitude and M the \textcolor{red}{Absolute magnitude}
\end{frame}

@@

\begin{frame}{Conclusions: forecast performances vs intepretability}
\begin{itemize}
\item The mgm algorithm is not the best one in term of performances, but it provides the connections between the features. On the other side, except for the variable importance in RF, the other are black box one
\item The mgm model, as the other graphical model is open to a true scientific validation, the other not. 
\item The probabilistic models lack in the forecast is definitely compensated by their interpretability 
\item This is meaningful since this two features are in conflict 
\item The probabilistic models provide a good trade-off between intepretability and forecast performances, as long as one is interest to produce a really scientific result (e.g if the only aim is the forecast the RF is definitely better. However how long one can trust to the RF result ?)
\end{itemize}
\end{frame}

@@

\section{Supporting Info}

\subsection{ROC and $\phi$ factor}
\begin{frame}{}
\begin{center}
{\Huge ROC and $\phi$ factor}
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\frametitle{$\phi$ coefficent}
\begin{table}[]
\begin{tabular}{c|c|c}
\multicolumn{1}{l|}{\textbf{}} & \textbf{Actual - N} & \textbf{Actual - P} \\ \hline
\textbf{Predicted - N}         & \#TP                & \#FN                \\ \hline
\textbf{Predicted - P}         & \#FP                & \#TP               
\end{tabular}
\end{table}
\begin{equation*}
\phi=\dfrac{TP\times TN - FP\times FN}{\sqrt{\left(TP+FP\right)\left(TP+FN\right)\left(TN+FP\right)\left(TN+FN\right)}}
\end{equation*}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Receiver operating characteristic}
\begin{columns}
\column{0.5\textwidth}
\begin{center}
\includegraphics[width=1\linewidth]{Pic/ROC/ROC_A.png}
\includegraphics[width=1\linewidth]{Pic/ROC/ROC_B.png}\\
\end{center}
\column{0.5\textwidth}
\begin{center}
\includegraphics[width=0.5\linewidth]{Pic/ROC/ROC_A_R.png}
\includegraphics[width=0.5\linewidth]{Pic/ROC/ROC_B_R.png}\\
\end{center}
\end{columns}
\begin{center}
Images taken from \cite{ROC_fig}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Receiver operating characteristic}
\begin{columns}
\column{0.5\textwidth}
\begin{center}
\includegraphics[width=1\linewidth]{Pic/ROC/ROC_C.png}
\includegraphics[width=1\linewidth]{Pic/ROC/ROC_D.png}\\
\end{center}
\column{0.5\textwidth}
\begin{center}
\includegraphics[width=0.5\linewidth]{Pic/ROC/ROC_C_R.png}
\includegraphics[width=0.5\linewidth]{Pic/ROC/ROC_D_R.png}\\
\end{center}
\end{columns}
\begin{center}
Images taken from \cite{ROC_fig}
\end{center}
\end{frame}

\subsection{FAMD}
\begin{frame}{}
\begin{center}
{\Huge FAMD}
\end{center}
\end{frame}

\begin{frame}{Factor analysis of mixed data - FAMD}
\begin{center}
r(z,k) correlation coefficient (z and k quantitative) \\
\end{center}
\begin{center}
$\eta^{2}$(z,q) correlation ratio (z quantitative and q qualitative)
\end{center}
\begin{center}
PCA $\rightarrow$ $max\sum_{k}r^{2}(z,k)$
\end{center}
\begin{center}
MCA $\rightarrow$ $max\sum_{q}\eta^{2}(z,q)$
\end{center}
\begin{center}
FAMD $\rightarrow$ $max\sum_{k}r^{2}(z,k)+max\sum_{q}\eta^{2}(z,q)$
\end{center}
\end{frame}

\subsection{Graphical models}
\begin{frame}{}
\begin{center}
{\Huge Graphical models}
\end{center}
\end{frame}

\begin{frame}{Graphical models}
\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
\begin{equation}
X  \perp Y| Z \iff  p(X,Y|Z) = p(X|Z)p(Y|Z)
\end{equation}
\begin{equation}
p(\textbf{x}_{1:V})=p(x_{1})\prod^{V}_{t=1}p(x_{t}|x_{t-1})
\end{equation}
\end{frame}

\begin{frame}{Graphical models}
\begin{theorem}[Hammersley-Clifford]
A positive distribution $p(\textbf{y})>0$ satisfies the CI properties of an indirect graph G iif p can be represented as a product of factor, one per maximal clique,  i.e.
\begin{equation}
p(\textbf{y}|\theta)= \dfrac{1}{Z(\theta)}\prod_{c \in C }\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
where C is the set of all the (maximal) cliques of G,  and Z($\theta$) is the partition function given by 
\begin{equation}
Z(\theta):= \sum_{y}\prod_{c\in C}\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
Note that this partition function is what ensures the overall distribution sums to 1
\end{theorem}
\end{frame}

\begin{frame}{Graphical models}
\begin{equation}
p(y|\theta)=\dfrac{1}{Z(\theta)} exp\left(-\sum_{c}E(y_{c}|\theta_{c})\right)
\end{equation}
\begin{equation}
\psi_{c}(y_{c}|\theta_{c})=exp\left(-E(y_{c}|\theta_{c})\right)
\end{equation}
\end{frame}

\begin{frame}{Parameters tuning}
\begin{equation}
\frac{\partial\mathcal{L}}{\partial\boldsymbol{\theta}_{c}}=\frac{1}{N}\sum_{i}\left[\phi_{c}(y_{i})-\frac{\partial}{\partial\boldsymbol{\theta}_{c}}\log Z(\boldsymbol{\theta})\right]
\end{equation}

\begin{equation}
\begin{split}
\frac{\partial \log Z(\boldsymbol{\theta})}{\partial\boldsymbol{\theta}}=\mathbb{E}\left[\phi_{c}(\textbf{y})\right|\theta]= \sum_{\textbf{y}}\phi_{c}(\textbf{y})p(\textbf{y}|\boldsymbol{\theta})
\end{split}
\end{equation}
\begin{equation}
\frac{\partial\mathcal{L}}{\partial\boldsymbol{\theta}_{c}}=\left[\frac{1}{N}\sum_{i}\phi_{c}(y_{i})\right]-\mathbb{E}\left[\phi_{c}(\textbf{y})\right]
\end{equation}
\begin{equation}
\frac{\partial\mathcal{L}}{\partial\boldsymbol{\theta}_{c}}=\mathbb{E}_{p_{emp}}\left[\phi_{c}(\textbf{y})\right]-\mathbb{E}_{p_{(\cdot|\boldsymbol{\theta})}}\left[\phi_{c}(\textbf{y})\right]
\end{equation}
\begin{equation}
\mathbb{E}_{p_{emp}}\left[\phi_{c}(\textbf{y})\right]=\mathbb{E}_{p_{(\cdot|\boldsymbol{\theta})}}\left[\phi_{c}(\textbf{y})\right]
\end{equation}
\end{frame}

\begin{frame}{Parameters tuning}
\begin{equation}
p\left(\textbf{y}|\boldsymbol{\theta}\right) = \dfrac{1}{Z(\theta)}\exp \left( \sum_{c}\boldsymbol{\theta}^{T}_{c}\phi_{c}\left(\textbf{y}\right)\right)
\end{equation}
\begin{equation}
\mathcal{L}\left(\boldsymbol{\theta}\right):= \frac{1}{N}\sum_{i}\log p\left(\textbf{y}_{i}|\boldsymbol{\theta}\right)=\frac{1}{N}\sum_{i}\left[\sum_{c} \boldsymbol{\theta}^{T}_{c}\phi_{c}(y_{i})-\log Z\left(\boldsymbol{\theta}\right)\right]
\end{equation}
\end{frame}

\begin{frame}{Gaussian graphical model}
\begin{equation}
f_{Y}(y)=\dfrac{det(K)^{1/2}}{(2\pi)^{d/2}}\exp\left(-\dfrac{1}{2}\textbf{y}^{T}K\textbf{y}\right)
\end{equation}

\begin{equation}
K=\Sigma^{-1}
\end{equation}

\begin{frame}{GLASSO}
\begin{equation}
L_{pen}(K,\hat{\mu})=log\det(K)-tr(K\;S)-\rho||K|| 
\end{equation}
\begin{center}
$K=\Sigma^{-1}$
\end{center}
\begin{center}
S: empirical covariance matrix
\end{center}
\end{frame}

\begin{frame}{mgm algorithm}
\begin{equation}
P(X_{s}|X_{\backslash s})=\exp\left\lbrace E_{s}(X_{\backslash s})\phi_{s}\left(X_{s}\right)+B_{s}(X_{s})-\Phi\left(X_{\backslash s}\right) \right\rbrace
\end{equation}
\begin{center}
$\phi_{s}$ function of sufficient statistics $B_{s}$ base measure
\end{center}

\begin{equation}
\begin{split}
P(X)= & \exp \left( \sum_{s\in V}\theta_{s}\phi_{s}(X_{s})+\sum_{s\in V}\sum_{r\in N(s)}\theta_{s,r}\phi_{s}(X_{s})\phi_{r}(X_{r})\\ +&...+\sum_{r_{1},...,r_{k}\in C}\theta_{r1,...,rk}\prod_{j=1}^{k}\phi_{r}_{j}\left(X_{r}_{j}\right)+\sum_{s\in V}B_{s}(X_{S})-\Phi(\theta) \right)
\end{split}
\end{equation}

\begin{equation}
\hat{\theta}=arg \min_{theta}\left\lbrace -\mathcal{L}(\theta,X)+\lambda ||\theta ||_{1}  \right\rbrace\quad ||\theta ||_{1}=\sum_{j=1}^{J}|\theta_{j}|
\end{equation}

\begin{frame}{mmod - Homogeneus Mixed Interaction (HMI) models}
\begin{center}
N observation, d discrete variables and q continuous variable. The observation has the form $x=(i,y)=(i_{1},...,i_{d},y_{1}...y_{q})$. The probability of discrete variables falling in the cell i is denoted as $p(i)$. The conditional distribution of continous variables to fall in the cell $i$ is given by the multivariate gaussian $\mathcal{N}(\mu(i),\Sigma)$
\end{center}

\begin{equation}
f(\textbf{i},\textbf{y})= p(\textbf{i})(2\pi)^{-q/2}det(\Sigma)^{-1/2}\exp\left[-\dfrac{1}{2}\left(\textbf{y}-\mu(\textbf{i})\right)^{T}\Sigma^{-1}\left(\textbf{y}-\mu(\textbf{i})\right)\right]
\label{gaussMix}
\end{equation}

\begin{equation}
f(\textbf{i},\textbf{y})=\exp\left[g(\textbf{i}) +h(\textbf{i})^{T}\textbf{y}-\frac{1}{2}y^{T}K\textbf{y}\right]
\end{equation}

\begin{center}
where $K$ is the concentration matrix, $g(i)$ and $h(i)$ are the log-linear expansion of the probability $p(\textbf{i})$ (canonical parameters)
\end{center}

\end{frame}
\begin{frame}{mmod}
\begin{center}
Decomposable model
\end{center}
\begin{equation}
\hat{f}(x)=\prod_{j=1}^{k}\dfrac{\hat{f}_{C_{j}}(x_{C}_{j})}{\hat{f}_{S_{j}}(x_{S}_{j})} 
\end{equation}
Maximized likehood
\begin{equation}
\hat{\mathcal{L}_{s}}=\sum_{i}n(i)log\left(\dfrac{n(i)}{N}\right)-Nq\dfrac{log(2\pi)}{2}-N\dfrac{log(det(S))}{2}-N\frac{q}{2}
\end{equation}

\begin{equation}
\hat{p}(\textbf{i})=n(\textbf{i})/N \quad \hat{\mu}(\textbf{i})=\bar{y}(\textbf{i})\quad\hat{\Sigma}=S=\sum_{i}n(i)S_{i}/N
\end{equation}

\subsection{Information theory}
\begin{frame}{}
\begin{center}
{\Huge Information theory}
\end{center}
\end{frame}

\begin{frame}{Information theory}
\begin{equation}
H(X)=-\sum_{x \in X} p(x)log p(x)
\end{equation}
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)} \quad D(p||q)\geq 0
\end{equation}

\begin{frame}{Information theory}
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
\begin{equation}
g(\alpha,\textbf{C},\textbf{S},f_{i})=MI(f_{i};\textbf{C})-\sum_{f_{s}\in S}\alpha(f_{i},f_{s},\textbf{C},\textbf{S})MI(f_{i};f_{s})
\end{equation}
\end{frame}

>>>>

% Uncomment these lines for an automatically generated outline.
%\begin{frame}{Outline}
%\setcounter{tocdepth}{1}
%\begin{center}
%  \tableofcontents
%\end{center}
%\end{frame}

@@

@@

\begin{frame}{Conclusions: forecast performances vs intepretability}
\begin{itemize}
\item The mgm algorithm is not the best one in terms of performance, but it provides the connections between the features. On the other side, except for the variable importance in RF, the SVM and RF are a black box one
\item The mgm model, as the different graphical model is open to a proper scientific validation, the SVM and the RF not. 
\item The probabilistic models lack in the forecast is compensated by their interpretability 
\item This is meaningful since interpretability and completeness/accuracy are in conflict 
\item It is worth noting that the logistic model, which is highly interpretable, has performances similar to the mgm
\item The probabilistic models provide a good trade-off between interpretability and forecast performances. As long as one is interested in producing a really scientific result, such algorithms as the logistic regression should be considered (e.g if the only aim is the forecast, the RF is better. However, how long one can trust the RF result ?)
\end{itemize}
\end{frame}

@@
